# Chapter 2: Central Tendency in R {.unnumbered}

This section demonstrates how to calculate and interpret the mean, median, and mode in R using data from the Programme for International Student Assessment (PISA). Additionally, we will explore skewness on the distribution of math achievement scores. All examples below use R code that can be adapted for your own datasets. 

To illustrate the calculations, let’s extract plausible value 1 math achievement (PV1MATH) from the PISA U.S. dataset. 

```{r}
#| echo: true
#| message: false
#| warning: false
library(haven) #to read .sav files 
data <- read_sav("chapter2/US_Data_22.sav") 
 
math_achi <- data$PV1MATH 
head(math_achi) #display the first few values 

```
## 1	  Calculating the Mean {.unnumbered}

```{r}
#| echo: true
#| message: false
#| warning: false
mean_math_achi <- mean(math_achi) 
mean_math_achi 

```

## 2    Calculating the Median {.unnumbered}
```{r}
#| echo: true
#| message: false
#| warning: false
median_math_achi <- median(math_achi) 
median_math_achi 
```
## 3    Calculating the Mode {.unnumbered}
Since R does not have a built-in function for mode, we will define a custom function. 
```{r}
#| echo: true
#| message: false
#| warning: false
# Custom function to calculate mode 
calculate_mode <- function(x) { 
  uniq_x <- unique(x) |> sort() 
  uniq_x[table(x) |> which.max()] 
} 
 
# Calculate the mode 
#the data are rounded to the nearest whole number to calculate the mode 
mode_math_achi <- calculate_mode(math_achi |> round()) 
mode_math_achi 
```
## 4 	  Examining Skewness {.unnumbered}
```{r}
#| echo: true
#| message: false
#| warning: false
library(psych) #for skew function 
skewness_math_achi <- skew(math_achi, na.rm = TRUE) 
skewness_math_achi 
```
```{r}
#| echo: true
#| message: false
#| warning: false
# Visualize the distribution 
hist(math_achi, main = "Distribution of Math Scores", xlab = "Math Scores",  
     col = "lightblue", border = "white") 
```
Recall that: 

- a positive skewness value indicates a right-skewed distribution (longer tail on the right), which is consistent with the histogram above. 

- when the data are right-skewed, the mean is greater than the median, and the mode is less than the median. These are consistent with what we observed in the calculations above. 