# Appendix {.unnumbered}

## Standard Normal Cumulative Probability Table {.unnumbered}

::: {.table-scroll}
```{r}
#| echo: false
#| results: asis
#| message: false
#| warning: false

library(readr)
library(knitr)
library(kableExtra)
library(dplyr)

z_table <- read_csv("tables/z_table.csv", show_col_types = FALSE)

z_table <- z_table |>
  mutate(across(-1, ~ sprintf("%.4f", .x)))
header_rows <- which(z_table[[1]] == "z")

tbl <- kbl(z_table, format = "html", align = "c") |>
  kable_styling(full_width = FALSE)

for (r in header_rows) {
  tbl <- row_spec(tbl, r, bold = TRUE)
}

tbl
```

:::

[Download full table (CSV)](tables/z_table.csv)



## Critical Values of $t$ Distribution Table {.unnumbered}
::: {.table-scroll}
```{r}
#| echo: false
#| results: asis
#| message: false
#| warning: false
t_table <- read_csv("tables/t_two_tail.csv", show_col_types = FALSE)

kbl(
  t_table,
  col.names = c("df", ".10", ".05", ".02", ".01"),
  align = "c",
  caption = "Critical Values of t Distribution"
) |>
  kable_styling(full_width = FALSE)
```
:::

*This table is organized by two-tail probability. To find the corresponding one-tail probability, divide the column probability by 2. To convert from one-tail probability to two-tail probability, multiply by 2.*

[Download full table (CSV)](tables/t_critical_values.csv.csv)

##  Critical Values of the Pearson ($r$) Correlation {.unnumbered}
::: {.table-scroll}
```{r}
#| echo: false
#| message: false
#| warning: false

library(readr)
library(knitr)
library(kableExtra)

r_table <- read_csv("tables/r_crit_val.csv", show_col_types = FALSE)

kbl(
  r_table,
  col.names = c("df", ".10", ".05", ".02", ".01"),
  align = "c",
  caption = "Critical Values of the Pearson (r) Correlation"
) |>
  kable_styling(full_width = FALSE)
```

:::
*This table is organized by two-tail probability. To obtain the corresponding one-tail probability, divide the column probability by 2. To convert from one-tail probability to two-tail probability, multiply by 2.*


[Download full table (CSV)](tables/r_crit_val.csv)


##  Critical Values of the Spearman Rho Correlation {.unnumbered}
::: {.table-scroll}
```{r}
#| echo: false
#| message: false
#| warning: false
rho_table <- read_csv("tables/spearman_crit.csv", show_col_types = FALSE)

kbl(
  rho_table,
  col.names = c("n", ".10", ".05", ".02", ".01"),
  align = "c",
  caption = "Critical Values of the Spearman (ρ) Correlation"
) |>
  kable_styling(
    full_width = FALSE,
    bootstrap_options = "basic"
  )
```
:::
*This table is organized by two-tail probability. To obtain the corresponding one-tail probability, divide the column probability by 2. To convert from one-tail probability to two-tail probability, multiply by 2.*


[Download full table (CSV)](tables/spearman_crit.csv)

## Critical Values of Chi-Square ($\chi^2$) {.unnumbered}

::: {.table-scroll}
```{r}
#| echo: false
#| message: false
#| warning: false

chisq_table <- read_csv("tables/chi_sq.csv", show_col_types = FALSE)

kbl(
  chisq_table,
  col.names = c("df", ".05", ".01", ".001"),
  align = "c",
  caption = "Critical Values of Chi-Square (χ²)"
) |>
  kable_styling(
    full_width = FALSE,
    bootstrap_options = "basic"
  )
```
:::
[Download full table (CSV)](tables/chi_sq.csv)


##  Critical Values for the Mann-Whitney U Test {.unnumbered}

::: {.table-scroll}

```{r}
#| echo: false
#| message: false
#| warning: false

mw_table <- read_csv("tables/mann_whitney_u.csv", show_col_types = FALSE)

kbl(
  mw_table,
  col.names = c("n₁ / n₂", "2","3","4","5","6","7","8","9","10","11","12","13","14","15"),
  align = "c",
  caption = "Critical Values for the Mann–Whitney U Test"
) |>
  kable_styling(
    full_width = FALSE,
    bootstrap_options = "basic"
  )
```

:::
*Unlike most other tests learned in this book, reject the null hypothesis if the test value is less than or equal to the value given in the table below.*

[Download full table (CSV)](mann_whitney_u.csv)


## Critical Values for the Wilcoxon Matched-Pairs Signed-Ranks Test {.unnumbered}

::: {.table-scroll}
```{r}
#| echo: false
#| message: false
#| warning: false


wilcox_table <- read_csv("tables/wilcoxon.csv", show_col_types = FALSE)

kbl(
  wilcox_table,
  col.names = c("n", ".05", ".01", ".001"),
  align = "c",
  caption = "Critical Values for the Wilcoxon Matched-Pairs Signed-Ranks Test"
) |>
  kable_styling(
    full_width = FALSE,
    bootstrap_options = "basic"
  )
```
:::
*Unlike most tests, reject the null hypothesis if the test statistic is less than or equal to the value in the table.*
*This table is organized by two-tail probability. To obtain the corresponding one-tail probability, divide the column probability by 2*

[Download full table (CSV)](wilcoxon.csv)

## Summary of Nonparametric Tests {.unnumbered}

::: {.formula-table}

| Test / Method | When to use? | Degrees of Freedom | Test Statistic | How to Evaluate |
|---------------|-------------|------------------|----------------|-----------------|
| **Chi-Square Goodness-of-Fit** | Categorical data; comparing one distribution to a known distribution | k − 1 | χ² | If obtained χ² is greater than the critical value (for given df), reject the null. Report Fisher’s Exact Test if available. |
| **Chi-Square Test for Independence** | Categorical data; testing for association between two variables | (k₁ − 1)(k₂ − 1) | χ² | If obtained χ² is greater than the critical value (for given df), reject the null. Report Fisher’s Exact Test if available. |
| **Independent Samples Chi-Square Test** | Categorical data; testing for association between two variables using summary data | (k₁ − 1)(k₂ − 1) | χ² | If obtained χ² is greater than the critical value (for given df), reject the null. Report Fisher’s Exact Test if available. |
| **Wilcoxon Signed Rank Test** | Nonparametric alternative to related-samples t test; converts measured values to ordinal data | n/a | T (smaller absolute rank sum) | Compare to critical value in Wilcoxon table. If T ≤ critical value, reject the null. |
| **Mann–Whitney U Test** | Nonparametric alternative to two independent samples t test; converts measured values to ordinal data | n/a | Uₓ or Uᵧ (smaller value) | Compare to critical value in Mann–Whitney table. If U ≤ critical value, reject the null. |
| **Kruskal–Wallis H Test** | Nonparametric alternative to one-way ANOVA; converts measured values to ordinal data | k − 1 | H | If obtained H is greater than the critical value (from chi-square table for given df), reject the null. |

:::