## Hypothesis Testing in R 

In this section, we will walk through the implementation of hypothesis testing in R, using the **one-sample z test** as introduced in this chapter. Our empirical example uses the PISA U.S. math achievement scores (PV1MATH) and compares them against the OECD mean of 500 points, as reported by OECD. Note that the OECD scaling assumes approximately normal distributions with a mean around 500 and a standard deviation of about 100. 

```{r}
#| echo: true
#| message: false
#| warning: false
library(haven) 
data <- read_sav("chapter6/chapter6data.sav") 
```

###   Manual Implementation of the One-Sample z Test 

We will follow the four key steps of hypothesis testing: 

####	State the Hypotheses

Our research question is: 

*Is the average math performance of U.S. students ($\mu$) significantly different from the OECD mean of 500? *

The hypotheses are stated as: 

- $H_0:\,\mu=500$
- $H_a:\,\mu\ne500$

This is a **two-tailed test**, since we are checking whether U.S. students score either higher or lower than the OECD mean. 
```{r}
#| echo: true
#| message: false
#| warning: false
# Compute sample statistics 
mean_us <- mean(data$PV1MATH, na.rm = TRUE) # mean of PV1MATH, excluding NAs 
n_us <- sum(!is.na(data$PV1MATH)) #sample size, excluding NAs 
 
mean_us; n_us 
```
####	Set the Decision Criteria 

We adopt the conventional significance level: 

- $\alpha=0.05$
- Corresponding critical $z$-values: $\pm1.96$. 

If the $p$ value < 0.05 (or equivalently, the test statistic falls beyond $\pm1.96$), we will reject $H_0$.

####	Compute the Test Statistic 

Review Chapters 4 and 5 for normality assumption checking. Given our data meet the assumptions, we move on to compute the z statistic. 
```{r}
#| echo: true
#| message: false
#| warning: false
# Parameters 
mu_oecd <- 500      # OECD population mean 
sigma_oecd <- 100   # OECD population standard deviation 
 
# Standard error 
sem <- sigma_oecd / sqrt(n_us) 
 
# z statistic 
z_value <- (mean_us - mu_oecd) / sem 
z_value 
```
z = -25.72 indicates that the U.S. mean is 25.72 standard deviations below the OECD mean. 

####	Compute the p Value and Draw a Conclusion 
```{r}
#| echo: true
#| message: false
#| warning: false
# p value for two-tailed test 
p_value <- 2 * (1 - pnorm(abs(z_value))) #1-pnorm(abs(z_value)) gives the upper tail probability  
p_value 
```
As $p<0.05$, we reject the null hypothesis and conclude that U.S. students’ average math score is significantly different from the OECD mean of 500. 

###    Using R’s Built-in Functions
```{r}
#| echo: true
#| message: false
#| warning: false
library(BSDA) 
# two-tailed test: H0: mu = 500 vs. Ha: mu != 500 
z_out <- z.test(x = data$PV1MATH, 
                mu = 500, 
                sigma.x = 100,          # OECD-population SD (assumed known) 
                alternative = "two.sided") 
z_out 
```
The output confirms our manual calculations, with a $z$ statistic of -25.72 and a $p$ value close to 0, leading to the same conclusion: we reject the null hypothesis. 
