# Analyzing Differences Between Two Groups  {.unnumbered}
 

## Learning Objectives {.unnumbered}

 

By the end of this chapter, you will be able to: 

- Identify and differentiate between different types of tests to compare two groups, understanding their appropriate applications based on the nature of the data and comparisons being made. 

- Understand differences between parametric and nonparametric tests.  

- Explain the logic behind conducting independent - and related-samples t tests on continuous variables, including verifying assumptions, calculating the test statistic, and interpreting results within the context of hypothesis testing. 

- Apply the Mann-Whitney $U$ test and Wilcoxon’s Matched-Pairs Signed-Ranks Test, demonstrating an understanding of their logic, appropriate use cases, and result interpretation. 

- Evaluate the practical significance of statistical results using measures such as Cohen's $d$ and eta-squared, providing context for the findings of t test analyses. 

In the previous chapter, we explored the one-sample $t$ test, a method used to compare a single group's mean to a known population mean. This gave us the foundation for understanding how $t$ tests work and their role in hypothesis testing. Now, we are moving on to explore additional types of $t$ tests to compare two groups. Before diving into the details of the independent-samples $t$ test, which is commonly used in educational research, let’s take a brief look at three types of $t$ tests. These include the one-sample $t$ test (covered in chapter 8), two independent-samples $t$ test, and two related-samples $t$ test, each tailored to different types of comparisons. Understanding the distinctions between these tests will help clarify when and how to use them effectively. 

## Types of $t$ Tests 

All three types of t tests involve comparing the values of a variable from one group of individuals to something else – and that *something else* tells you what kind of t test you should use. To explain the three tests, imagine that you have a class of third grade students, and you have a list of each student’s reading achievement scores. The following table explains how you might analyze those scores with each type of t test, depending on what you’re comparing them to.  

::: {.formula-table }

| What to Compare | T-Test Used | Example |
|---|---|---|
| **One single number** | One-sample *t* test | You want to see if this group’s scores are, on average, significantly different from the state average for reading achievement scores. |
| **A different group’s scores on the same variable** | Two-independent-samples *t* test | You want to compare this group’s reading achievement scores to those from another third-grade class. |
| **The same group’s scores on a different variable<br>or at a different time** | Two related-samples *t* test | You want to compare this group’s reading scores to their math achievement scores, to see if their achievement levels are different across the two subjects.<br><br>**OR**<br><br>You want to compare this group’s reading scores after third grade to the same group’s reading scores after fourth grade, to see if they change over time. |

:::


Now, we will provide a very brief overview of each type.  

 

### One-sample t test 

 

This should be very familiar since we just learned it in the last chapter. You use this t test if you know the population mean ($\mu$), but you do not know the population standard deviation ($\sigma$). You are trying to find out if your sample mean ($\bar{X}$) is significantly different from a known population mean. To run this analysis, we will use only one continuous variable or column of data – each individual in the group has one value or score for the variable of interest. 

 

### Two-independent-samples t test 


Commonly in practice, you do not know a population mean to use as a comparison point. You are instead comparing sets of numbers obtained from two separate groups in order to draw inferences about their populations. Maybe you have two groups of students, one of which received some kind of intervention, and you want to compare end-of-year achievement scores between them. You might want to know if female college students have higher GPAs than male students. This time, your dataset will need two columns to run the test: one to indicate which group the individual is in, and one with their value for the dependent variable. 
 

### Related-samples t test 
 

You are still comparing two sets of numbers, but this time both sets belong to the same people or two groups that are somewhat related. For example, you might be comparing achievement scores at the beginning and end of a year to see if a class improved. You might also compare outcomes under two different conditions, such as how many times students were absent when class met in-person compared to the number of absences for the same students when the school shifted to remote learning. Instead of having the same group of people, you sometimes have a matched group (e.g., siblings or a set of individuals who have been matched based on their scores on other variables). Either way, your dataset will look different yet again. Like with the one-sample t test, you only have a single list of individuals or a single set of matched pairs. However, this time they each have two columns of information (e.g., pre-test and post-test scores, or English scores and math scores). 


We hope this helps you avoid getting confused about the three different types of t tests. The details for how to conduct each analysis are similar but there are important differences. The names are actually pretty good clues: 

- *One*-sample: you only have one set of data (e.g., a set of reading scores from a sample of third graders), so you need a population value (e.g., the national average of reading score) for comparison. 

- Two *independent*-samples: you have two distinct groups that are not related or connected in any way – one might even say they are two independent groups. Membership in a group is mutually exclusive, meaning that no individual or case can belong to both groups simultaneously.  

- Two *related*-samples (also known as paired-sample or repeated-measures test): here you have two sets of numbers that are connected – either the same people at different times or different conditions, or people who are paired together.  

 

And now we will dive into the details of the independent-samples t test. This is probably one of the most used methods to compare two groups in educational research. 

## Overview of the Two-Independent-Samples t Test 

 

As we did when learning about the one-sample t test, let’s again walk through an example from educational research. Similar to the previous example, the teacher is now concerned if her class average of English Language Arts (ELA) scores ($\bar{X} = 425.26, SD = 44.99$) is significantly different from another class of third graders whose teacher joined the pilot implementation of a new reading intervention. When the result was returned the class average of the fellow teacher was 440 with a SD of 35. Are these two averages significantly different? Let’s find out.  

```{r}
#| label: fig-9.1
#| fig-cap: "Figure 9.1. Overlapping histograms of ELA scores for two third-grade classes"
#| fig-width: 8
#| fig-height: 4.6
#| warning: false
#| message: false
#| echo: false

library(ggplot2)
library(scales)

set.seed(4)
group1 <- rnorm(25, mean = 425.26, sd = 44.99)
group2 <- rnorm(25, mean = 440, sd = 35)

df <- rbind(
  data.frame(score = group1, group = "Group 1"),
  data.frame(score = group2, group = "Group 2")
)

m1 <- mean(group1)
m2 <- mean(group2)

# bins similar to your screenshot
binwidth <- 15

ggplot(df, aes(x = score, fill = group)) +
  geom_histogram(
    binwidth = binwidth,
    position = "identity",
    alpha = 0.45,
    color = NA
  ) +
  geom_vline(xintercept = m1, linetype = "dashed", linewidth = 1, color = "blue") +
  geom_vline(xintercept = m2, linetype = "dashed", linewidth = 1, color = "red") +
  annotate(
    "label",
    x = m1, y = 1.2,
    label = paste0("Group 1 Mean = ", round(m1, 2)),
    size = 3,
    label.size = 0.25,
    fill = "white",
    color = "blue"
  ) +
  annotate(
    "segment",
    x = m1, xend = m1, y = 1.0, yend = 0.1,
    linewidth = 0.6, color = "blue",
    arrow = arrow(length = unit(0.12, "inches"))
  ) +
  annotate(
    "label",
    x = m2, y = 1.2,
    label = paste0("Group 2 Mean = ", round(m2, 0)),
    size = 3,
    label.size = 0.25,
    fill = "white",
    color = "red"
  ) +
  annotate(
    "segment",
    x = m2, xend = m2, y = 1.0, yend = 0.1,
    linewidth = 0.6, color = "red",
    arrow = arrow(length = unit(0.12, "inches"))
  ) +
  scale_fill_manual(
    values = c("Group 1" = "#8fbcd4", "Group 2" = "#e89a9a"),
    breaks = c("Group 1", "Group 2")
  ) +
  coord_cartesian(xlim = c(300, 525)) +
  labs(
    x = "ELA Score",
    y = "Frequency",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = c(0.9, 0.9),
    legend.background = element_rect(fill = "white", color = "grey80"),
    panel.grid.minor = element_blank()
  )
```

Similar to the previous one-sample t test example, we will complete four steps of hypothesis testing.  

### Step 1: Determine the null and alternative hypotheses. {.unnumbered}

 

With a one-sample test, the null hypothesis states there is no difference between the sample mean and the population mean. We are now comparing the means of two independent samples, to determine statistically if their population means are identical or different. So, the null hypothesis about population means is that there is no difference between the population means of the two groups. The way we indicate is that one population mean ($\mu_1$ for Group 1) minus the other ($\mu_2$ for Group 2) is zero.  

$$
H_0: \mu_1− \mu_2=0
$$
 
 

Therefore, the alternative hypothesis is that there is a difference between the two groups: 

 

$$
H_A: \mu_1− \mu_2\ne 0
$$
 

 

Note that this is the alternative hypothesis for a two-sided or two-tailed test. We are not assuming that one mean is larger than the other, just that they are different. If you had reason to believe that one group mean would be larger than the other, your alternative hypothesis would be: 

$$
H_A: \mu_1 > \mu_2
$$
 

or 

 
$$
H_A: \mu_1 < \mu_2
$$
 

### Step 2: Set the criteria for a decision. {.unnumbered}
 

The next step in testing the null hypothesis is to set the criteria for a decision: how much of a difference is significant enough that we can conclude it is not only due to random chance? We decide that $\alpha = 0.05$ as usual, but the degrees of freedom are slightly different for an independent-samples t test compared to a one-sample t test. This time, we must subtract one from each group, so $df = (n_1 − 1)+(n_2 − 1) = n_1 + n_2−2 = N - 2$, where 
$n_1$ and $n_2$ refer to the sample size of Group 1 and Group 2, respectively, and $N$ refers to the total number of subjects in both groups combined.  
 

Flip to the Critical Values of $t$ Distribution Table at the back of the book like before, looking for the intersection between your alpha level and your degrees of freedom to find the critical value.  For our example, both groups have a sample size of $35$ so the $df  = 70 - 2 = 68$. Since we don’t have $df = 68$, round down to $50$ df to be conservative. The critical $t$ value associated with 50 df at $\alpha = 0.05$ for a two-tailed test is $\pm2.009$.  

```{r}
#| label: fig-9.2
#| fig-cap: "t-distribution with two-tailed rejection regions"
#| fig-width: 8
#| fig-height: 4.6
#| warning: false
#| message: false
#| echo: false

library(ggplot2)


df <- 49
alpha <- 0.05
tcrit <- qt(1 - alpha/2, df)

x <- seq(-4, 4, length.out = 2000)
curve_df <- data.frame(x = x, y = dt(x, df))

ggplot(curve_df, aes(x, y)) +

  # T distribution line
  geom_line(aes(color = "T-Distribution"), linewidth = 1) +

  # Rejection regions (tails only)
  geom_area(
    data = subset(curve_df, x <= -tcrit),
    aes(fill = "Rejection Region"),
    alpha = 0.3
  ) +
  geom_area(
    data = subset(curve_df, x >= tcrit),
    aes(fill = "Rejection Region"),
    alpha = 0.3
  ) +

  # Critical value lines
  geom_vline(xintercept = -tcrit, linetype = "dashed", color = "red") +
  geom_vline(xintercept = tcrit,  linetype = "dashed", color = "red") +

  # Null line
  geom_vline(
    aes(linetype = "Null Hypothesis (Mean Diff = 0)"),
    xintercept = 0,
    color = "black"
  ) +

  scale_color_manual(values = c("T-Distribution" = "blue"), name = NULL) +
  scale_fill_manual(values = c("Rejection Region" = "red"), name = NULL) +
  scale_linetype_manual(
    values = c("Null Hypothesis (Mean Diff = 0)" = "dashed"),
    name = NULL
  ) +

  labs(
    x = "Standardized Difference in Sample Means (t)",
    y = "Density"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    legend.position = c(0.72, 0.62),
    legend.background = element_rect(fill = "white", color = "grey80")
  )
```

### Step 3: Check assumptions and compute the test statistic.  {.unnumbered}

 

The third step has a few parts. First, we verify that our data meet the assumptions required for this method of analysis, and then we calculate the test statistic. The assumptions for an independent-samples t test are:  

- Normality: The dependent variable is approximately normally distributed (as before, this requirement is less important if n > 30).  

- Random sampling: You are using a random sample of independent observations.  

- Independence: The scores are independent of one another. 

- Equal variances: The variances of the dependent variable are equal across two groups or populations being compared.   

 

Equal variance is new, but it is important to keep in mind for this method. Remember earlier in the book when we said means are important and so is standard deviation (or variance, which is the same basic concept)? Here is one place where that comes into play. Imagine you are comparing scores of two groups on the same test. In one group, all scores are tightly clustered near the mean. In the other group, they are completely spread out with a much wider range than compared to the first group. It is harder to make a solid conclusion about the differences between two group means when the distributions are completely different.  

 
```{r}
#| label: fig-9.3
#| fig-cap: "Normal distributions with equal vs. unequal variances"
#| fig-width: 8
#| fig-height: 4.6
#| warning: false
#| message: false
#| echo: false

library(ggplot2)

x_min <- 0
x_max <- 125
x <- seq(x_min, x_max, length.out = 2000)

params <- data.frame(
  scenario = c("Equal Variances (X: 0-125)", "Equal Variances (X: 0-125)",
               "Unequal Variances (X: 0-125)", "Unequal Variances (X: 0-125)"),
  group    = c("Group 1", "Group 2", "Group 1", "Group 2"),
  mean     = c(50, 60, 50, 60),
  sd       = c(10, 10, 10, 20),
  color    = c("blue", "darkgreen", "blue", "darkgreen")
)

# Build curve data
curve_df <- do.call(rbind, lapply(seq_len(nrow(params)), function(i){
  p <- params[i, ]
  data.frame(
    x = x,
    y = dnorm(x, mean = p$mean, sd = p$sd),
    scenario = p$scenario,
    group = sprintf("%s (Mean=%d, SD=%d)", p$group, p$mean, p$sd),
    mean = p$mean,
    color = p$color
  )
}))

# Means for vlines/legend
means_df <- unique(curve_df[, c("scenario", "group", "mean", "color")])

ggplot(curve_df, aes(x = x, y = y, color = group)) +
  geom_line(linewidth = 1) +
  geom_vline(
    data = means_df,
    aes(xintercept = mean, color = group),
    linetype = "dashed",
    linewidth = 0.8,
    show.legend = FALSE
  ) +
  facet_wrap(~scenario, nrow = 1) +
  coord_cartesian(xlim = c(x_min, x_max)) +
  labs(x = "Scores", y = NULL, color = NULL) +
  scale_color_manual(
    values = setNames(means_df$color, means_df$group)
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )

```
When analyzing data from two groups, it is essential to compute and evaluate the size of variance for each group. This helps in understanding the spread of the data and determining if the variances are similar or different. Drawing histograms for each group is particularly useful for visually inspecting the data spread and identifying any noticeable differences between two distributions.  

#### Levene's Test for Assessing Equal Variances {.unnumbered}

**Levene's Test** is a statistical procedure used to assess whether the variances of a dependent variable are equal across different groups. This test is typically performed before conducting analyses that assume equal variances, such as t tests or ANOVA, to determine if this assumption holds. We won’t show you how to calculate Levene’s Test by hand because statistical software packages will do this for you.  
 

However, the important thing to remember is that if the result of Levene’s Test is statistically significant $(p < .05)$, your variances are *not equal*. If your variances are not equal, you should consider using a different test that we will learn later or, at minimum, report the statistical results adjusted for unequal variances between two groups, which is provided by most software packages as well. If your Levene’s Test results are not statistically significant $(p > .05)$, your group variances are considered equal.  

### Step 4a: Find the p value.  {.unnumbered}

Next, we calculate the test statistic. We are going to show you the formula first, but please don’t panic. It is not as bad as it seems.  

$$
t = \frac{(\bar{X}_1-\bar{X}_2)-(\mu_1-\mu_2)}{\sqrt{s_p^2(\frac{1}{n_1}+\frac{1}{n_2})}}
$$

Let’s take that a little at a time, starting with the numerator. The first part is the difference between group means in your sample. Then we are comparing that to the expected difference between group means in the population (i.e., the mean differences specified in the null hypothesis). Look back at the null hypothesis for this type of test. What is the expected difference between group means in the population? Zero! For practical purposes, the top half of the fraction is just the difference between group means.  

 

Do you remember what was in the denominator of the t statistic for a one-sample t test? It was the standard error, or the sample standard deviation divided by the square root of the sample size. This denominator is more complicated because we have two separate groups. Under the null hypothesis, we assume that both sample groups came from the same population, but most likely the standard errors are not exactly the same. So which sample variance do we use as an estimate for the population variance? The optimal solution is to make full use of the variance estimates from both samples to obtain a more accurate estimate of the population variance.  

There is another new value in the denominator, $s_p^2$
,
 
 which represents the pooled sample variance. The formula for this looks very complicated, but it is just a way of weighting the two sample variances by the sample size. If one group is bigger than the other, their sample variance is weighted more heavily by this formula. If you notice this looks a little different from what you might find elsewhere, that is because we are being more specific about how you get to the degrees of freedom, with n – 1.   

$$
s_p^2 = \frac{(n_1 − 1)s_1^2 + (n_2 − 1)s_2^2}{(n_1+ n_2)−2}
$$
 
 

What does all that mean? You are multiplying the sample variance for group one by the degrees of freedom for group one and adding that to the same thing for group two. Then you are dividing the outcome by the degrees of freedom for the whole test. In our example, the sample size of each group is 35. Thus,   

$$
s_p^2 = \frac{(34)s_1^2 + (34)s_2^2}{68}
$$
 
Now you have a measure of sample variance that you can use for the whole combined sample of both groups. If both groups have the same sample size, like our example, there is a much simpler version of the formula you can use.  

$$
s_p^2 = \frac{s_1^2 + s_2^2}{2}
$$

In our example, $s_1^2 = (44.99)^2$, and $s_2^2 = (35)^2$. Let’s plug in these values and compute the $s_p^2$

$$
s_p^2 = \frac{s_1^2 + s_2^2}{2} = \frac{(44.99)^2 + (35)^2}{2} = 1624.55
$$

Either way, once you have the value for $s_p^2$, you bring that back to the $t$ statistic formula and do what we always do: standard deviation divided by the square root of the sample size. It just looks more complicated when we have two separate sample sizes, and because this formula starts with sample variance and then takes the square root, rather than starting with standard deviation, which is already the square root of the variance. 

$$
t = \frac{(\bar{X}_1-\bar{X}_2)-(\mu_1-\mu_2)}{\sqrt{s_p^2(\frac{1}{n_1}+\frac{1}{n_2})}}
= \frac{(425.26−440)−0}{\sqrt{1624.55(\frac{1}{35}+\frac{1}{35})}} 
= \frac{-14.74}{9.634}
= -1.53
$$

 
If you are completely lost, try not to get too caught up in the math and focus on the concept. We are finding the difference between the two group means and dividing it by the standard error for the difference. It is just more complicated to calculate this version of standard error than it has been for other tests. The good news is that statistical software programs will do the math for you.  

```{r}
#| label: fig-9.4
#| fig-cap: "Two-tailed t-test with critical values"
#| fig-width: 8
#| fig-height: 4.6
#| warning: false
#| message: false
#| echo: false

library(ggplot2)

df <- 49
tcrit <- 2.009
t_stat <- -1.53

# Curve data
x <- seq(-4, 4, length.out = 2000)
curve_df <- data.frame(
  x = x,
  y = dt(x, df)
)

# Tail shading data
left_tail  <- subset(curve_df, x <= -tcrit)
right_tail <- subset(curve_df, x >=  tcrit)

ggplot(curve_df, aes(x, y)) +

  # t distribution curve
  geom_line(aes(color = "t-distribution"), linewidth = 1) +

  # rejection regions
  geom_area(data = left_tail,
            aes(fill = "Rejection Region"),
            alpha = 0.3) +
  geom_area(data = right_tail,
            aes(fill = "Rejection Region"),
            alpha = 0.3) +

  # critical value lines
  geom_vline(aes(linetype = "Critical values: ±2.009"),
             xintercept = -tcrit,
             color = "black",
             linewidth = 0.9) +
  geom_vline(aes(linetype = "Critical values: ±2.009"),
             xintercept = tcrit,
             color = "black",
             linewidth = 0.9) +

  # test statistic line
  geom_vline(aes(color = "Test Statistic"),
             xintercept = t_stat,
             linewidth = 1) +

  # label for test statistic
  annotate("label",
           x = t_stat,
           y = 0.02,
           label = "-1.53",
           size = 3,
           label.size = 0.25,
           fill = "white") +

  scale_color_manual(
    values = c(
      "t-distribution" = "blue",
      "Test Statistic" = "darkgreen"
    ),
    name = NULL
  ) +
  scale_fill_manual(
    values = c("Rejection Region" = "red"),
    name = NULL
  ) +
  scale_linetype_manual(
    values = c("Critical values: ±2.009" = "dashed"),
    name = NULL
  ) +

  labs(
    x = "t-value",
    y = "Density"
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +

  coord_cartesian(xlim = c(-4, 4)) +

  theme_minimal(base_size = 12) +
  theme(
    legend.position = c(0.78, 0.80),
    legend.background = element_rect(fill = "white", color = "grey80"),
    panel.grid.minor = element_blank()
  )
```

The last step in testing a hypothesis is more straightforward: compare the value of the $t$ statistic you just calculated to the critical value you identified in step-two. If the $t$ statistic is larger, meaning farther out from the center of the distribution in either direction, then a significant difference exists between the groups.  
 

In this example, the observed $t$ value is $= -1.53$ which is greater than the lower critical $t$ value of $-2.009$. Therefore, we fail to reject the null hypothesis. The actual $p$ value for $t = -1.53$ in a two-tail test is $0.131$.  

### Step 4b: Draw a conclusion and report your conclusion.  {.unnumbered} 

Since we failed to reject the null hypothesis, which states that there is no difference in population means of the two groups, we conclude that the ELA classroom averages taught by two different teachers are not significantly different. The observed mean difference between two groups is simply by chance due to sampling. Note that we do not calculate the effect size to assess practical significance, as no statistically significant difference was found. 
 

How might we report the result in APA style? An independent-samples t test was conducted to compare ELA classroom averages between students taught by two different teachers. The results indicated that there was no statistically significant difference in classroom averages between the two groups, $t(68) = 1.53, p = .131$, suggesting that the mean ELA classroom scores for students in these two classrooms are comparable. 

## Effect Size for the Independent-Samples $t$ Test 

If you find a significant difference, you should calculate the effect size. We most commonly report effect size for this method using estimated Cohen’s d. If you remember back to chapter 8, Cohen's d quantifies the magnitude of the difference between two means relative to the variability in the data. Similarly, we can use Cohen’s d as a standardized measure of practical significance, helping to interpret the importance of the result. However, the formula for Cohen’s d that you use for an independent-sample t test is slightly different from what we have already learned: 

$$
d=\frac{M_1-M_2}{\sqrt{s_p^2}}
$$

The numerator is just the difference in sample means, and the denominator is the pooled sample standard deviation; you can find the formula earlier in the chapter.  

Eta-squared ($\eta^2$) is another measure of effect size used to quantify the proportion of variance in a dependent variable that is associated with one or more independent variables in an analysis. It provides an estimate of how much of the variability in the dependent variable is explained by the grouping variable. Eta-squared can be calculated using the following formula:

$$
\eta^2=\frac{t^2}{t^2+df}
$$

Where: 

- $t^2$ is the square of the t test statistic. 

- $df$  is the degrees of freedom for the test. 

Either way, you only report effect size if you find a significant result when conducting your hypothesis test. 



## Overview of the Related-Samples $t$ Test 

We have finally arrived at the last type of t test, and to make things extra interesting, it goes by several different names depending on how you are using it.  

**Related-samples $t$ test** is typically used as a general term for a $t$ test that involves comparisons of related or dependent groups. Additionally, it is used when you have the same participants measured on two dependent variables that use the same or comparable scales, such as math vs reading tests. For example, a researcher may examine whether there is a significant difference between students’ performance on standardized tests in two subjects.  

Two specific types of the related-samples $t$ test are often used in research, each corresponding to a particular research design:   

- **Repeated-measures $t$ test** is used when we measure one group of participants twice on the same or comparative measure at different points in time. In educational research, this test is often applied in pre/post design, where the goal of the analysis is to assess change in scores before and after an intervention, or at the start and end of an academic year. For example, a study might measure students’ reading comprehension before and after a literacy intervention to evaluate its effectiveness.  

- **Paired-samples $t$ test** is used when individuals or cases are matched in pairs based on key characteristics before comparing their scores. Unlike repeated-measures design, the paired individuals are not the same person measured twice; rather they are two different people who are statistically equivalent as pairs except for the dependent variables. For example, you could statistically match students who experience an intervention with a control group of students who are very similar on key variables. Other times family members (siblings, married couples) are paired, if it is reasonable to assume they are similar in important ways. This type of t test is less common as it often requires more advanced statistical methods or poses logistical challenges in finding well-matched pairs. 
 
The related-samples t test is actually a special case of the one-sample t test. In a related-samples test, you collect data in pairs, and the difference between each pair is calculated. Then, we compute statistics based on these differences. Your beginning assumption, or null hypothesis, is that there will be no difference between the pairs. You could create a new variable in your dataset that was equivalent to the difference between pairs and use a one-sample t test to compare that difference variable to the expected value of 0.  

 

Let’s assume we are not creating new variables, and we will show you how to directly conduct a related-samples t test. However, keep in mind that you would get the same results either way.  

 

We know it can be hard to understand why a related-samples t test is not the same as an independent-samples t test, since we are interested in the differences between two sets of scores in both cases. It really matters statistically that the same, or similar, people are in both groups in a related-samples study. The short description is that when you have different people in your two groups, it is hard to know for sure if the differences are really due to your grouping factor (e.g., free/reduced lunch status, treatment/control for an intervention), or if you by chance ended up with people who were very different from each other for reasons that had nothing to do with your grouping factor.  

 

With a related-samples t test, we don’t need to worry about any of that. We know there are not random differences between the people in our two groups because the same people are in both groups. From a statistical perspective, this changes two things: we calculate degrees of freedom differently, and our hypotheses and test statistics are based on the difference between scores. Let’s use an example to look at each step of the related-samples t test more closely.  

 

Imagine that you developed a new student learning intervention and piloted it with a sample of nine randomly selected students. You have data on their scores before and after the intervention, and now you want to determine if student scores significantly changed. Your dataset would look like this: 

| Pre | Post | Difference |
|:---:|:----:|:----------:|
| 10  | 14   | 4          |
| 10  | 9    | -1         |
| 11  | 16   | 5          |
| 13  | 14   | 1          |
| 14  | 17   | 3          |
| 15  | 18   | 3          |
| 16  | 18   | 2          |
| 16  | 17   | 1          |
| 17  | 20   | 3          |

If this was an independent-samples t test, you would be thinking about the distribution of *pre* scores and the distribution of post scores. However, in a related-samples test, the focus shifts to the distribution of *difference* scores (i.e., post – pre), which in this example ranges from -1 to 5 and has a sample mean of 2.33 with a SD of 1.80.  For this example, the research question we will address is whether scores changed significantly between pre and post, which may suggest an effect of the intervention.   

### Step 1: Determine the null and alternative hypotheses. {.unnumbered}

Recall that with an independent-samples t test, there are two separate sample means to consider, so the null hypothesis for a two-tailed test is:  

$$
H_0: \mu_1 − \mu_2 = 0
$$
 


With a related-samples t test, our null hypothesis for a two-tailed test is: 

 
$$
H_0: \mu_D = 0
$$


The alternative hypothesis for a two-tailed, non-directional test is: 

 
$$
H_A: \mu_D \ne 0
$$
 

The null hypothesis states that the difference between two dependent variables is zero and the alternative is that the difference between dependent variables is not zero. Keep in mind you can also conduct a one-tailed, directional test by making the alternative hypothesis that 
$\mu_D < 0$ or $\mu_D > 0$

### Step 2: Set the criteria for a decision. {.unnumbered}

 

The next step in testing the null hypothesis is to set the criteria for a decision: how much of a difference is significant enough that we can conclude it is not a result of chance alone? As usual, we can decide that $\alpha = .05$. The degrees of freedom for a related-samples t test are simply calculated by taking n – 1, where n refers to the number of pairs. Since each participant contributes one pair of scores (before and after), the number of pairs is equal to the number of participants in the study (with no missing values on both tests). You can use a t table just as before, looking for the intersection between your alpha level and your degrees of freedom to find the critical value. In our example, because n = 9,  df = 9 - 1 = 8. If you flip to the Critical Values of t Distribution Table and look at df = 8 and $\alpha = .05$ (two-tailed), you see the critical value for this test is $\pm2.306$. 

```{r}
#| label: fig-9.5
#| fig-width: 8
#| fig-height: 5
#| fig-align: center
#| fig-cap: "Sampling distribution with two-tailed rejection regions"
#| warning: false
#| message: false
#| echo: false

library(ggplot2)

df <- 8
alpha <- 0.05
tcrit <- qt(1 - alpha/2, df)

x <- seq(-4, 4, length.out = 1000)

data <- data.frame(
  x = x,
  y = dt(x, df)
)

ggplot(data, aes(x, y)) +

  # t distribution curve
  geom_line(color = "blue", linewidth = 1.2) +

  # rejection regions
  geom_area(
    data = subset(data, x <= -tcrit),
    fill = "red",
    alpha = .5
  ) +

  geom_area(
    data = subset(data, x >= tcrit),
    fill = "red",
    alpha = .5
  ) +

  # critical value lines
  geom_vline(xintercept = c(-tcrit, tcrit),
             linetype = "dashed",
             linewidth = 1) +

  scale_y_continuous(
    expand = c(0, 0)   # removes gap between shaded region and axis
  ) +

  labs(
    x = "t-value",
    y = NULL
  ) +

  annotate(
    "text",
    x = 2.7,
    y = .35,
    label = paste0("Critical values: ±", round(tcrit,2)),
    size = 4
  ) +

  theme_minimal(base_size = 14) +

  theme(
    axis.title.y = element_blank(),
    panel.grid = element_blank()
  )
```

### Step 3: Check assumptions and compute the test statistic.  {.unnumbered}

 

The third step has a few parts. First, we verify that our data meet the assumptions required for this method of analysis, and then we calculate the test statistic. The assumptions for a related-samples t test are:  

- Normality: The distribution of the difference scores is approximately normal. As with other t tests, this assumption is less critical when the sample size is greater than 30 due to the central limit theorem. 

- Random sampling: You are using a random sample of independent observations.  

- Independence: The scores are independent of one another. Although the two groups in a related-samples t test are obviously dependent on each other – since they consist of paired scores from the same individuals or matched pairs – the scores within each group must still be independent of one another. This means that each participant’s score should not be influenced by another participant’s score within the same group. In other words, the assumption of independence applies to the individual observations within each group, not between groups. 

Notice that we don’t have the equal variances assumption as we did with the independent-samples t test. We have that assumption with independent-samples t tests because we are concerned that major differences in group distributions mean the groups are very different, but as we said before, since we are using the same people this time, we are not as worried about that.   



### Step 4a: Find the p value.  {.unnumbered}

Next, we calculate the test statistic. You will notice a lot of Ds in the following formula because we are *focusing on difference scores*, not two separate sample means. Instead of having two distributions of sample means, we have one *distribution of sample differences*.  

$$
t = \frac{\bar{D} - \mu_D}{\frac{S_D}{\sqrt{n}}}
$$

Let’s unpack each part of the formula separately to make things easier to digest. The numerator is looking for the difference between the sample mean difference and the expected difference in the population under the assumption that the null hypothesis is true. However, recall that our null hypothesis is that there is no difference in the population, so $\mu_D$ is zero. For our purposes, therefore, the numerator is just the sample mean difference. Where does that come from? It is calculated as the average of the individual difference scores (post – pre) across all pairs in the sample, which is 2.33 in our example.  
 

Now we look at the denominator of the test statistic formula. This should look familiar by now because we see a standard deviation divided by the square root of the sample size (i.e., standard error). The first value is $S_D$, which is something we have not seen before. That represents the standard deviation of the difference distribution, which in this case is 1.80. Conveniently, our sample size is 9, so the square root of n is 3. Using this, we calculate the standard error of the sampling distribution of difference scores is 0.6. Therefore, we can calculate the test statistic as follows: 

$$
t = \frac{2.33 - 0}{\frac{1.80}{3}} = \frac{2.33}{0.6} = 3.883
$$

The next step in testing our hypothesis is to compare our observed test statistic with the critical value. Our test statistic is definitely farther out into the tails of the distribution than 2.306, so we can reject the null hypothesis.  

```{r}
#| label: fig-9.6
#| fig-width: 8
#| fig-height: 5
#| fig-align: center
#| fig-cap: "Sampling distribution with rejection regions and calculated t-value."
#| warning: false
#| message: false
#| echo: false

library(ggplot2)

df <- 8
alpha <- 0.05
tcrit <- qt(1 - alpha/2, df)
t_calc <- 3.88

x <- seq(-4, 4.2, length.out = 1000)

data <- data.frame(
  x = x,
  y = dt(x, df)
)

ggplot(data, aes(x, y)) +

  geom_line(color = "blue", linewidth = 1.2) +

  geom_area(
    data = subset(data, x <= -tcrit),
    fill = "red",
    alpha = .5
  ) +

  geom_area(
    data = subset(data, x >= tcrit),
    fill = "red",
    alpha = .5
  ) +

  geom_vline(
    xintercept = c(-tcrit, tcrit),
    linetype = "dashed",
    linewidth = 1
  ) +

  geom_vline(
    xintercept = t_calc,
    color = "darkgreen",
    linetype = "dashed",
    linewidth = 1
  ) +

  annotate(
    "segment",
    x = t_calc,
    xend = t_calc,
    y = 0.02,
    yend = 0,
    colour = "darkgreen",
    arrow = arrow(length = unit(0.15,"cm"))
  ) +

  annotate(
    "label",
    x = t_calc + 0.25,
    y = 0.03,
    label = "3.88",
    size = 4
  ) +

  scale_y_continuous(
    expand = c(0,0)
  ) +

  labs(
    x = "t-value",
    y = NULL
  ) +

  theme_minimal(base_size = 14) +

  theme(
    axis.title.y = element_blank(),
    panel.grid = element_blank()
  )
```

The actual $p$ value associated with the observed t statistic of 3.88 with $df = 8$ for a two-tailed test is 0.0047, which is smaller than 0.05.  

 

### Step 4b: Draw a conclusion and report your conclusion.   {.unnumbered}

 

The small $p$ value ($p = 0.0047$) indicates that, under the assumption that the null hypothesis is true, the probability of observing a difference as extreme as, or more extreme than, the one we found in the sample is 0.47%. Because this is highly unlikely, we reject the null hypothesis and conclude that there is a statistically significant difference in scores between pre and post. Because the observed difference is statistically significant, we now want to know how the difference is practically meaningful by calculating the effect size.  

 

## Effect Size for Related-Samples $t$ Tests 

 

Once we have found a statistically significant difference, it is useful to examine the practical significance as well. The formula for Cohen’s $d$ for this type of t test is as follows:  

$$
d = \frac{M_d}{S_D}
$$

That is, we take the mean difference and divide it by the standard deviation of the difference. The effect size for the example is calculated as:  

$$
d = \frac{M_d}{S_D} = \frac{2.33}{1.8} = 1.29
$$

This means that, on average, scores increased by 1.29 standard deviation units from pre and post, which represents a large effect size. Based on this statistical result, and assuming no other confounding factors, we may conclude that the intervention had a meaningful impact on scores!   

 

Here is how you would report this example using APA style: A repeated-measures t test was conducted to evaluate the effect of the intervention on scores using a random sample of nine students. Results indicated that, on average, students’ scores changed by 2.33 points between pre- and post-tests. The difference was statistically significant, $t(8) = 3.88, \, p < .01$. Cohen’s $d = 1.29$, which is a large effect size, indicating substantial improvement in scores following the intervention.  


There is one extra wrinkle to consider: *what if the expected difference in the population is not zero?* For example, K-12 students take standardized tests in math and reading multiple times each year, and their scores are generally expected to improve over time. Testing companies use population data to provide expected growth benchmarks for each grade level.    

 

Suppose third graders are expected to see a 13-point increase in math scores from the beginning to the end of the school year. If you wanted to determine whether students in your school were experiencing growth that was different from the expected population growth, your null and alternative hypotheses would be as follows: 

$$
H_0 : \mu_D = 13
$$

$$
H_A : \mu_D \ne 13
$$

## Comparing Two Groups with Ordinal Data: Nonparametric Tests 

In hypothesis testing, parametric and nonparametric tests differ primarily in their assumptions about the data. In general, parametric tests, such as t tests, are distinguished by the assumption of normality on the data with a continuous dependent variable. Parametric tests typically use the raw data directly in their calculations and require certain conditions, such as homogeneity of variance and interval or ratio-level measurement of the dependent variable.  


In contrast, nonparametric tests are distinguished by the absence of any distributional requirement for the data and are less restrictive in terms of the scales of measurement. These tests are more flexible and can be used with ordinal data or when assumptions for parametric tests are violated. Instead of analyzing raw values, nonparametric methods often involve rank-based transformations of the data before analysis, making them less sensitive to outliers and skewed distributions. 
 

A difference between the two approaches is that parametric tests allow for inferences about population parameters, such as means and standard deviations. However, nonparametric tests do not estimate specific population parameters; they instead assess whether two or more groups share the same distribution. The primary advantage of nonparametric tests is their robustness and fewer assumptions, making them useful alternatives to $t$ tests when data are not normally distributed, have small sample sizes, or contain outliers.


## Choosing between Parametric and Nonparametric Tests 

In general, we prefer to use the test that is the most efficient compared to others and produces the largest statistical power among all alternatives. For example, consider the two independent samples t test. If the scores are independent and homoscedastic, and follow a normal distribution, this test is optimal in that there is no other test with greater statistical power for a given alternative. On the other hand, failure to satisfy the normality assumption means that the t test may not be optimal. In particular, the power of nonparametric competitors often exceeds that of the t test when the population distribution is symmetric but is flatter (or heavy-tailed), when the distribution is highly skewed, or if the distribution is multimodal and heavy-tailed.  

 

If data assumptions of parametric tests are not adequately satisfied, the actual α of a test (i.e., the probability of having a Type I error) may differ substantially from the user-specified α level (e.g., 0.05), although we never know exactly how much the probability differs. Especially when sample size exceeds 30, there is evidence that many parametric tests, especially those comparing means, maintain control over Type I error rates even when the dependent variable is not normally distributed. If the distribution is skewed and/or multimodal and heavy-tailed, nonparametric tests frequently do a better job of controlling for Type I error rates.  

 

In educational research, nonparametric tests are particularly useful when the outcome variable includes responses on a Likert-type rating scale, because the distribution is often skewed. You may come across research that uses parametric tests, such as a two-independent samples t test, to compare responses on a single Likert scale item between two groups. However, this practice is not generally recommended. Additionally, as we have discussed, outliers can contribute a great deal of error variance in parametric tests, so a nonparametric alternative potentially is more powerful.  

 

The following table shows parametric tests you are already familiar with and nonparametric alternatives. We will describe the Mann-Whitney U test and the Wilcoxon signed rank T test in more detail.  

::: {.striped-table}

| **Parametric** | **Nonparametric** |
|---|---|
| One-sample *t* test | Sign test |
| Independent-samples *t* test | Mann–Whitney *U* test |
| Related-samples *t* test | Wilcoxon signed rank *T* test |
| One-way between-subjects ANOVA | Kruskal–Wallis *H* test |
| One-way repeated measures ANOVA | Friedman test |

:::

## Two Independent-Sample Mann-Whitney $U$ Test 

 

**The Mann-Whitney $U$ test** is a statistical procedure used to determine whether the dispersion of ranks is equal across two independent groups. It is used as a nonparametric alternative to the two independent-samples t test. The test has three assumptions: 

- Both samples are randomly drawn from their respective populations. 

- In addition to independence within each sample, there is mutual independence between the two samples (no one belongs to both groups). 

- The measurement scale is at least ordinal (no categorical variables). 

 

Let’s walk through an example. Suppose that we give nine children a math achievement test and record their scores. They then receive daily practice on arithmetic problems. At the end of each practice session, their papers are marked by a teacher’s aide, who praises them for the number they got right and states that they might do even better in the next few days. We treat eight other children the same except for the statements made by the teacher’s aide. For these children, they point out the problems they missed and tell them that they should be able to do better than that. At the end of two weeks, we retest all children on a parallel form of the math achievement test and calculate their “gain scores” over the two-week period: 

::: {.score-table}

| Criticism (X) | -3 | -2 | 0 | 0 | 2 | 5 | 7 | 9 | |
|---------------|----|----|---|---|---|---|---|---|---|
| Praise (Y)    | 0  | 2  | 3 | 4 | 10 | 12 | 14 | 19 | 21 |

:::

$$
H_0 \text{: The two groups have identical distributions.} 
$$
$$
H_A: \text{The two groups do not have identical distributions.} 
$$
 

These data are measured on a continuous scale, so theoretically we could use a two independent-samples t test, but first we should look at the descriptive statistics to see if our data meet the assumptions of that method. You can look at descriptive statistics for each group, but we think the most helpful way to consider the data is by viewing histograms for each group:  

```{r}
#| label: fig-9.7
#| fig-width: 7.5
#| fig-height: 4.5
#| fig-align: center
#| warning: false
#| message: false
#| echo: false

library(ggplot2)

criticism <- c(-3, -2, 0, 0, 2, 5, 7, 9)
praise <- c(0, 2, 3, 4, 10, 12, 14, 19, 21)

df <- rbind(
  data.frame(GainScore = criticism, Group = "Criticism"),
  data.frame(GainScore = praise, Group = "Praise")
)

ggplot(df, aes(x = GainScore, fill = Group)) +
  geom_histogram(
    binwidth = 2,
    position = "identity",
    alpha = 0.5,
    color = "black",
    boundary = 0
  ) +
  scale_x_continuous(
    name = "Gain Score",
    breaks = seq(-6, 22, by = 5)
  ) +
  scale_y_continuous(
    name = "Frequency",
    breaks = c(1, 2),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("Criticism" = "blue", "Praise" = "pink")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Given the very small sample sizes (8 and 9 cases for Criticism and Praise, respectively) and two distributions not looking normal, we should not use the independent-samples $t$ test. Since this is a very small dataset, let’s calculate the test statistic by hand to illustrate the logic behind the test. In practice, statistical software can perform this analysis. 

 

**Step 1:** Label the two groups $X$ and $Y$. The smaller group should be labeled $X$; if sample sizes are equal then it doesn’t matter which group is labeled $X$.  


$X$ = Criticism condition 

$Y$ = Praise condition 

 

**Step 2:** Combine all the scores into one distribution of $n_X$ + $n_Y$ cases. Arrange them in order of the dependent variable from lowest to highest and assign ranks.  


| **Condition** | **Gain Scores** | **Assigned Rank** |
|:---------:|:-----------:|:-------------:|
| X | -3 | 1 |
| X | -2 | 2 |
| X | 0 | 4 |
| X | 0 | 4 |
| Y | 0 | 4 |
| X | 2 | 6.5 |
| Y | 2 | 6.5 |
| Y | 3 | 8 |
| Y | 4 | 9 |
| X | 5 | 10 |
| X | 7 | 11 |
| X |9 |12 |
| Y |10 |13 |
| Y |12 |14 |
| Y |14 |15 |
| Y |19 |16 |
| Y | 21 | 17|

When assigning ranks in a dataset, ties occur when two or more scores are identical. In such cases, instead of giving each tied score a unique rank, we calculate the average rank they would have occupied. 

 

For example, in the table above, three students had gain scores of 0. These scores would have been ranked 3, 4, and 5 if assigned individually. Instead, we sum these ranks (3 + 4 + 5 = 12) and divide by 3, giving each student a rank of 4. The next score is then assigned the rank that follows the tied ranks, which in this case is 6.5 (the average of the next two ranks, 6 and 7). 

 

**Step 3**: Divide the cases back into their original groups.  

::: {.rank-table}

| **Criticism (X)** | **Assigned Rank** | **Praise (Y)** | **Assigned Rank** |
|:------------:|:-------------:|:----------:|:-------------:|
| -3 | 1   | 0  | 4   |
| -2 | 2   | 2  | 6.5 |
| 0  | 4   | 3  | 8   |
| 0  | 4   | 4  | 9   |
| 2  | 6.5 | 10 | 13  |
| 5  | 10  | 12 | 14  |
| 7  | 11  | 14 | 15  |
| 9  | 12  | 19 | 16  |
|    |     | 21 | 17  |

:::


**Step 4**: Sum up the ranks for each group.  

 
$$
\text{Sum of ranks for Group X: } 
\sum R_X = 1 +2+4+4+6.5+10+11+12=50.5
$$

$$
\text{Sum of ranks for Group Y: }
\sum R_Y=4+6.5+8+9+13+14+15+16+17=102.5
$$

**Step 5**: Use the summed ranks to calculate the test statistic (i.e., Mann-Whitney $U$ Statistic).  

$$
U_x = n_x n_y + \frac{n_x(n_x + 1)}{2} - \sum R_x
= 8 \times 9 + \frac{8(9)}{2} - 50.5
= 57.5
$$

$$
U_y = n_x n_y + \frac{n_y(n_y + 1)}{2} - \sum R_y
= 8 \times 9 + \frac{9(10)}{2} - 102.5
= 14.5
$$

- $n_X$ and $n_Y$ are the sample sizes of groups X and Y, respectively.
- $R_X$ and $R_Y$ are the rank sums for each group.
- U is the smaller of $U_X$ and $U_Y$.

**Step 6**: Evaluate the test statistic using the Critical Values for the Mann-Whitney $U$ Test table in the appendix. We use the value of $U = 14.5$ as the test statistic because it is the smaller of the two U values calculated. 

 

Next, we determine whether we are conducting a one-tailed or two-tailed test. In this case, our test is non-directional (two-tailed), so we divide our alpha level of 0.05 equally between the two tails, resulting in 0.025 in each tail. 

 

We locate the critical value by finding the number of cases in each group along the left side and top of the table. For group sizes of 8 and 9, the critical value is 15. Since our test statistic ($U = 14.5$) is less than or equal to the critical value (15), we reject the null hypothesis that the two distributions are the same. 

 

Note that, unlike many parametric tests where we compare the test statistic to see if it exceeds the critical value, in the Mann-Whitney $U$ test, we reject the null hypothesis when the test statistic is less than or equal to the critical value. 

 

With this result, we conclude that the distributions of the two groups are statistically different, indicating that the scores between the Criticism and Praise groups differ significantly, $U = 14.5$, $p = 0.043$. 



## Wilcoxon’s Matched-Pairs Signed-Ranks Test 


While the Mann-Whitney $U$ Test is used to compare two independent groups, **Wilcoxon’s Matched-Pairs Signed-Ranks Test** is a nonparametric alternative to the related-samples (or repeated-measures) t test.  

 

Let’s walk through another example to help you learn more about this method. Suppose we are studying the effect of a new instructional strategy on math performance, specifically on the speed and accuracy of solving arithmetic problems. We form 10 pairs of students, matched according to their pre-test score earned on similar math problems, and then randomly assign members of each pair to either the new instructional strategy or the traditional teaching method (control group). After completing a structured learning session, the students take a post-test assessing their arithmetic skills. 

 

Their post-test scores were recorded in the table below as Traditional (control) and New Strategy (treatment). Although different students are assigned to the two groups, the pairing process creates dependence between their scores, as students were first matched by pre-test ability before assignment.  

 

As we are interested in any difference (positive or negative) due to the instructional strategy, we use a two-tailed test. The null and alternative hypotheses for this investigation are:  

$H_0$: There is no difference in student performance between two instructional groups. 

$H_A$: There is a difference in student performance between two instructional groups. 

Rather than create new tables for each of the steps required to calculate the test statistic by hand, we will add a row to the table below showing which columns belong to which step in the process.  

 

- **Step 1**. Recode the paired scores in two columns.  

- **Step 2**. Obtain the difference between members of each pair (X-Y). 

- **Step 3**. Disregard the sign of the difference obtained and then assign ranks to the absolute magnitude of the difference. This means that a difference of -8 would be a higher rank than a difference of 5. If any difference is zero, disregard it in subsequent calculations. For tied ranks, assign the mid rank as we did in the prior test.  

- **Step 4**. Reassign the appropriate sign to the rank. This is a weird step but don’t take it literally, e.g., -2 does not mean the rank of -2 (negative 2nd). It simply provides two types of information about the score at once: the rank of the score and the direction of the difference. 

- **Step 5**. Compute the summed rank for both positive and negative differences. The smaller absolute value summation becomes our test statistic (6.5). 

::: {.wilcoxon-table-header}
| Steps | 1 | 2 | 3 | 4 | 5|
|:---:|:---:|:---:|:---:|:---:|:---:|
:::
::: {.wilcoxon-table}

| Pair | Traditional (X) | New (Y) | Difference (X-Y) | Absolute Difference | Rank of Absolute Difference | Signed rank of X-Y | Positive Difference | Negative Difference |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| 1 | 24 | 28 | -4 | 4 | 2 | -2 |  | -2 |
| 2 | 39 | 29 | 10 | 10 | 6.5 | 6.5 | 6.5 |  |
| 3 | 29 | 34 | -5 | 5 | 3.5 | -3.5 |  | -3.5 |
| 4 | 28 | 21 | 7 | 7 | 5 | 5 | 5 |  |
| 5 | 25 | 28 | -3 | 3 | 1 | -1 |  | -1 |
| 6 | 32 | 15 | 17 | 17 | 10 | 10 | 10 |  |
| 7 | 31 | 17 | 14 | 14 | 8 | 8 | 8 |  |
| 8 | 33 | 28 | 5 | 5 | 3.5 | 3.5 | 3.5 |  |
| 9 | 31 | 16 | 15 | 15 | 9 | 9 | 9 |  |
| 10 | 22 | 12 | 10 | 10 | 6.5 | 6.5 | 6.5 |  |
| **SUM** |  |  |  |  | **55** |  | **48.5** | **-6.5** |

:::

We then use the Critical Values for the Wilcoxon Matched-Pairs Signed-Ranks Test table in the appendix. Since $n = 10$ and we are conducting a two-sided test with an alpha level of .05, the critical value is 8. Our test statistic is 6.5, and similar to the Mann-Whitney U test, we reject the null hypothesis if the test statistic is less than or equal to the critical value. In this case, since 6.5 ≤ 8, we reject the null hypothesis. 

 

Alternatively, statistical software can be used to obtain the associated p value for the test statistic. In this case, the computed $p$ value = 0.037, which is less than 0.05, further supporting the rejection of the null hypothesis. Therefore, we conclude that there is a statistically significant difference in scores between the two instructional strategies. This suggests that the new instructional strategy has an effect. However, it is important to note that the test itself only determines whether a difference exists; it does not indicate the direction of the effect (i.e., whether the new strategy increased or decreased scores). 



