# Predicting an Outcome in R {.unnumbered}


In this section, we will demonstrate how to conduct and interpret simple linear regression, multiple linear regression, and logistic regression using educational data. The examples use the following variables from the PISA dataset: 

- **PV1MATH**: Math achievement score (continuous, outcome variable for linear regression) 

- **MATHPERS**: Effort and Persistence in Mathematics (continuous, predictor for simple and multiple regression) 

- **GROSAGR**: Growth Mindset (continuous, additional predictor for multiple regression) 

- **MATHEASE**: Perception of Mathematics as easier than other subjects (binary (0=No/1-Yes), outcome for logistic regression) 

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: false
# Load required package 
library(haven)       # For reading SPSS files 
library(tidyverse) 
library(car)         # For VIF and diagnostics 
library(broom)       # For tidy model summaries 
library(psych)       # For describe 
library(performance) # For diagnostics 
library(ResourceSelection) # For logistic regression goodness-of-fit 
 
# Load the dataset 
data <- read_sav("chapter12/chapter12data.sav") 
 
# Clean/select variables 
data <- data |> select(PV1MATH, MATHPERS, GROSAGR, MATHEASE) 
```


## 1	Simple Linear Regression {.unnumbered}

**Research Question**: Does effort and persistence in mathematics (MATHPERS) predict math achievement (PV1MATH)? 

### 1.1	Fit Simple Linear Regression  {.unnumbered}

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: false

simple_lm <- lm(PV1MATH ~ MATHPERS, data = data) 
summary(simple_lm)
```

**Interpretation**: 

The regression analysis found a significant positive relationship: for each one point increase in effort and persistence, math achievement scores increased by approximately 16.13 points ($b=16.13,SE=1.52,t=10.61,p<.01$).The model explained about 2.8% of the variance in math achievement ($R^2 = .028$). The residual standard error was 94.06, indicating that, on average, students’ actual math scores deviated from the predicted scores by about 94 points. Although the effect size is modest, the result suggests that students who put in more effort and persist in mathematics tend to score higher on the PISA math achievement test. 

### 1.2	Assumptions Checking {.unnumbered}

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: false
# Linearity, Homoscedasticity, Normality, Outliers 
par(mfrow = c(2, 2)) 
plot(simple_lm) 
```

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: false
par(mfrow = c(1, 1)) 
# Additional: Independence is typically assumed in survey data 
```

**Interpretation**: 

- **Linearity & Homoscedasticity**: The Residuals vs Fitted (upper left) and Scale-Location (lower left) plots show that residuals are roughly evenly spread around zero across fitted values, without clear patterns or fanning, suggesting linearity and constant variance (homoscedasticity). 

- **Normality**: The Q-Q plot of residuals displays points closely following the diagonal line, indicating that residuals are approximately normally distributed. 

- **Influential Points**: The Residuals vs Leverage plot shows that most points have low leverage, with no point deviating significantly from the rest, suggesting no influential outliers. 

## 2	Multiple Linear Regression {.unnumbered}

For the multiple linear regression, we will include both effort and persistence and growth mindset as predictors of math achievement. 

**Research Question**: Do effort and persistence (MATHPERS) and growth mindset (GROSAGR) jointly predict math achievement (PV1MATH)? 

### 2.1	Fit Multiple Linear Regression {.unnumbered}
```{r}
#| echo: true
#| message: false
#| warning: false
#| include: false
multi_lm <- lm(PV1MATH ~ MATHPERS + GROSAGR, data = data) 
summary(multi_lm) 
```
**Interpretation**: 

(Check section Reporting Results of a Multiple Regression Analysis for APA-style sample write-up) 

The multiple linear regression show that both predictors have significant positive effects on math achievement: for each one point increase in effort and persistence, math scores increased by about 13.74 points ($b=13.74,SE=1.51,p<.01$), and for each one standard deviation increase in growth mindset, scores increased by about 23.65 points ($b=23.65,SE=1.83,p<.01 $). The model explained approximately 6.7% of the variance in math achievement scores (
$R^2=.067$), with a residual standard error of 92.17. 

Compared to the previous simple regression model that included only effort and persistence, adding growth mindset as a second predictor increased the proportion of explained variance from 2.8% to 6.7%. This suggests that considering both effort and growth mindset provides a better prediction of students’ math achievement. 

### 2.2	Multicollinearity & Assumptions Checking {.unnumbered}
```{r}
#| echo: true
#| message: false
#| warning: false
#| include: false
vif(multi_lm) 
```

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: false
par(mfrow = c(2, 2)) 
plot(multi_lm) 
```

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: false
par(mfrow = c(1, 1)) 
```
**Interpretation**: 

VIFs for both predictors are approximately 1.02, which is well below the common threshold of 5. This indicates that multicollinearity is not a concern in this model, and the two predictors are not highly correlated with each other. 

The diagnostic plots (as those for the simple linear regression) indicate that the assumptions of linearity, homoscedasticity, normality of residuals, and the absence of influential outliers are generally met for this multiple linear regression model. 


## 3	Logistic Regression {.unnumbered}

**Research Question**: Does math achievement (PV1MATH) predict whether a student perceives mathematics as easier than other subjects (MATHEASE)? 

### 3.1	Prepare Outcome Variable {.unnumbered}
```{r}
#| echo: true
#| message: false
#| warning: false
#| include: false
data$MATHEASE <- factor(data$MATHEASE, levels = c(0, 1), labels = c("No", "Yes")) 
data$MATHEASE |> table() 
```
Among the student sample, 3497 perceived mathematics as easier than other subjects, while 501 did not. 


### 3.2	Fit Logistic Regression {.unnumbered}

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: false
logit_model <- glm(MATHEASE ~ PV1MATH, data = data, family = "binomial") 
summary(logit_model)  
```

Interpretation: 

A logistic regression was conducted to examine whether students’ math scores predict the likelihood that they perceive mathematics as easier than other subjects. The results show a significant positive effect of math score ($b=0.0034, SE=0.0005,z=6.79,p<.01$). This means that for each point increase in the math score, the odds of perceiving math as easier increase by 0.34% (odds ratio: $e^{0.0034} = 1.0034$). In other words, higher math achievement is associated with a slightly higher probability of perceiving math as easier than other subjects. 
