# Analyzing Change and Adjusting for Covariates in R {.unnumbered}


In this section, we demonstrate how to implement two of the central methods from this Chapter in R: **Repeated-Measures ANOVA** and **Analysis of Covariance (ANCOVA)**. Both approaches help educational researchers make more nuanced conclusions: the first by analyzing within-subject change across multiple measures, the second by adjusting for covariates to clarify between-group differences. 

```{r}
#| echo: true
#| message: false
#| warning: false
library(haven) 
data <- read_sav("chapter13/chapter13data.sav") 
```

## 1	Repeated-Measures ANOVA {.unnumbered}

We will examine U.S. students’ math performance across three content areas from PISA: 

- PV1MPRE: Reasoning 

- PV1MCQN: Quantity 

- PV1MCSS: Space & Shape 

These serve as within-subject factors (each student is measured in all three domains). For a between-subjects factor, we will include MATHMOT - Motivation to Do Well in Mathematics, a binary categorical variable. A value of 1 indicates that the student reported being more motivated to do well in mathematics than in other subjects, and 0 indicates they are not. 

### 1.1	Preparing the data {.unnumbered}

```{r}
#| echo: true
#| message: false
#| warning: false
# Load required packages 
library(tidyverse) 
library(ez) 
 
# Select variables 
rm_data <- data %>% 
  select(ID = CNTSTUID, PV1MPRE, PV1MCQN, PV1MCSS, MATHMOT) #CNTSTUID - Student ID 
 
# Reshape to long format for ezANOVA 
rm_long <- rm_data %>% 
  pivot_longer(cols = c(PV1MPRE, PV1MCQN, PV1MCSS), 
               names_to = "ContentArea", 
               values_to = "Score") 
 
head(rm_long |> as.data.frame()) 
```

### 1.2	Running the repeated-measures ANOVA {.unnumbered}

```{r}
#| echo: true
#| message: false
#| warning: false
library(dplyr) 
 
# Repeated measures ANOVA with MATHMOT as between-subjects factor 
anova_res <- ezANOVA( 
  data = rm_long, 
  dv = Score,              # dependent variable 
  wid = ID,                # participant ID 
  within = .(ContentArea), # within-subject factor 
  between = .(MATHMOT),    # between-subject factor 
  type = 3,                # Type III SS, to address unbalanced design 
  detailed = TRUE          # returns extra information 
) 
 
# Extract the ANOVA table from ezANOVA 
anova_table <- anova_res$ANOVA 
 
# Compute partial eta squared for each effect 
anova_table$eta_p2 <- with(anova_table, SSn / (SSn + SSd)) 
anova_table_rounded <- anova_table |> 
  mutate(across(where(is.numeric), ~ round(.x, 3))) 
anova_table_rounded 
```

#### 1.2.1	Between-Subjects Factor: MATHMOT {.unnumbered}

- Students with higher motivation in mathematics (MATHMOT) had significantly different scores across the math content areas overall, $F(1, 4008) = 4.16, \, p = .041$. 

- The partial $\eta^2$ (eta_p2) is .001, indicating that although the effect is statistically significant, the effect size is extremely small (about 0.1% of the variance explained). This reveals that large samples can make very small effects look statistically significant. 

#### 1.2.2	Within-Subjects Factor: ContentArea {.unnumbered}

- There is a significant effect of Content Area on math scores, $F(2, 8016) = 123.92$, $p < .001$, indicating that students’ performance varies across the three content areas. 

- The partial $\eta^2$ is .030, suggesting that about 3% of the variance in scores can be attributed to differences between content areas. This is a larger but still modest effect size. 

#### 1.2.3	Interaction: MATHMOT × ContentArea {.unnumbered}

The interaction between MATHMOT and Content Area is not significant, $F(2, 8016) = 1.75, \, p= .174$. This means that although motivated students scored differently overall, their relative performance patterns across the three content areas were not significantly different from less motivated peers. In other words, motivation shifts the overall level but not the shape of the profile. 

#### 1.2.4	Assumption Check: Sphericity {.unnumbered}

- Mauchly’s Test ($W  = 0.983, \, p < .001$) is significant. This suggests the assumption of sphericity is formally violated: the variances of the differences between content areas are not perfectly equal. However, with such a large sample ($N > 4000$), Mauchly’s test is extremely sensitive — even tiny deviations from sphericity can become statistically significant. 

- As the chapter explained, we apply corrections. For the effect of content areas: Greenhouse-Geisser (p[GG]) $p < .001$; Huynh-Feldt (p[HF]) $p < .001$. Both lead to the same conclusion: the ContentArea effect is robustly significant, even after adjustment. 

### 1.3	Post Hoc Comparisons {.unnumbered}

#### 1.3.1	Visualize the Means {.unnumbered}

```{r}
#| echo: true
#| message: false
#| warning: false
# Mean plot by ContentArea 
library(ggplot2) 
 
means_plot <- rm_long %>% 
  group_by(ContentArea) %>% 
  summarise(mean_score = mean(Score, na.rm = TRUE), 
            se = sd(Score, na.rm = TRUE) / sqrt(n())) 
means_plot 
```

```{r}
#| echo: true
#| message: false
#| warning: false
# Mean plot (line plot) with error bars 
ggplot(means_plot, aes(x = ContentArea, y = mean_score, group = 1)) + 
  geom_line(color = "skyblue", size = 1) + 
  geom_point(size = 3, color = "darkblue") + 
  geom_errorbar(aes(ymin = mean_score - se, ymax = mean_score + se), 
                width = 0.1, color = "darkblue") + 
  scale_y_continuous(limits = c(445, 465)) +   # adjust y-axis range 
  scale_x_discrete(labels = c("PV1MPRE" = "Reasoning", 
                              "PV1MCQN" = "Quantity", 
                              "PV1MCSS" = "Space & Shape")) + 
  ylab("Mean Math Score") + 
  xlab("Content Area") + 
  theme_minimal() + 
  ggtitle("Mean Math Scores Across Content Areas") 
```

#### 1.3.2	Pairwise Comparisons (Repeated-Measures t-tests) {.unnumbered}

We use pairwise.t.test() with Bonferroni correction. Since the data are within-subjects, we need to specify paired = TRUE. 

```{r}
#| echo: true
#| message: false
#| warning: false
# Pairwise repeated-measures t-tests 
pairwise_results <- pairwise.t.test(rm_long$Score,  
                                    rm_long$ContentArea, 
                                    paired = TRUE, 
                                    p.adjust.method = "bonferroni") 
pairwise_results 
```

From the line plot, students scored lowest in Space & Shape (450), higher in Quantity (458), and highest in Reasoning (461). The pairwise t-tests confirm that all three pairwise differences are statistically significant after Bonferroni correction ($p < .001$ for all comparisons), although the largest gap is between Space & Shape and the other two areas. 



## 2	ANCOVA {.unnumbered}

Now we turn to ANCOVA， where the focus is where the focus is comparing the three types of schools on math performance, while controlling for growth mindset: 

- **DV**: PV1MATH (overall math achievement) 

- **IV**: SC014Q01TA (School type: 1 - religious, 2 - non-profit, 4 - government) 

- **Covariate**: GROSAGR (Growth Mindset) 

### 2.1	Running ANCOVA {.unnumbered}

```{r}
#| echo: true
#| message: false
#| warning: false
library(emmeans) 
```

```{r}
#| echo: true
#| message: false
#| warning: false
# convert School Type to factor 
data$SchoolType <- factor(data$SC014Q01TA, 
                          levels = c(1,2,4), 
                          labels = c("Religious", "Non-profit", "Government")) 
data$SchoolType |> table() # check distribution 
```

```{r}
#| echo: true
#| message: false
#| warning: false
# ANCOVA model 
ancova_model <- aov(PV1MATH ~ SchoolType + GROSAGR, data = data) 
summary(ancova_model) 
```

```{r}
#| echo: true
#| message: false
#| warning: false
#Post-hoc Comparisons of adjusted means for School Type 
emmeans(ancova_model, pairwise ~ SchoolType) 
```

### 2.1.1	School Type Effect {.unnumbered}

- After controlling for growth mindset, there are still significant differences in math scores between school types, $F(2,3816)=41.68, \, p<.001$. This means that even if students had the same level of growth mindset, school type is still associated with different mean math achievement. 

- From the post-hoc comparison, after controlling for growth mindset (i.e., adjusted means): 
    
    - Religious schools ($M = 473$) scored significantly higher than Non-profit schools ($M = 405$), $p < .001$. 

    - Government schools ($M = 469$) also scored significantly higher than Non-profit schools, $p < .001$. 

    - There was no significant difference between Religious and Government schools (mean difference $= 3.82, p = .93$). 

Check the One-Way ANOVA results from Chapter 10 to see how these comparisons differ before controlling for growth mindset. 


#### 2.1.2	Growth Mindset Effect (Covariate) {.unnumbered}

Growth mindset has a significant effect on math achievement, $F(1,3816) = 188.65, \, p < .001$. 


### 2.2	Checking assumptions {.unnumbered}

#### 2.2.1	Independence of covariate and IV {.unnumbered}

Recall that one key assumption of ANCOVA is that the covariate (growth mindset) is associated with the dependent variable (math achievement) but not the independent variable (school type). We have shown that growth had a significant effect on math achievement. Now we check whether growth mindset differs by school type. 

```{r}
#| echo: true
#| message: false
#| warning: false
# Does SchoolType predict Growth Mindset? 
anova_cov <- aov(GROSAGR ~ SchoolType, data=data) 
summary(anova_cov) 
```

The ANOVA showed that growth mindset scores did not significantly differ across school types, $F(2,3817) = 1.60, \, p = .203$. This indicates that the covariate (growth mindset) is independent of the independent variable (school type), upholding the assumption for ANCOVA. 

#### 2.2.2	Homogeneity of regression slopes {.unnumbered}

Another assumption of ANCOVA is no interaction between the effect of the covariate and group membership. We test this by adding an interaction term between school type and growth mindset. 

```{r}
#| echo: true
#| message: false
#| warning: false
# Add interaction term 
interaction_model <- aov(PV1MATH ~ SchoolType * GROSAGR, data=data) 
#note. including the interaction term automatically includes the main effects 
summary(interaction_model) 
```

The interaction between school type and growth mindset was not significant, $F(2,3814) = 1.67, \, p = .189$, indicating that the effect of growth mindset on math achievement is consistent across school types. This upholds the homogeneity of regression slopes assumption for ANCOVA. 
